Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2015-05-10 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Added test and refs to Intro - draft ack

\dedication{To my wife and family}  work family

\abstract{
Uncontrolled fires result in loss of life, property and can create an adverse economic impact on a community.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus reduce the losses from fires.  However, as the result of the review of several arson cases and a review by the U.S. National Academy of Sciences, questions about the lack of the use of science in the practice of fire investigation were raised.  Specifically the National Academy of Sciences indicated that ``... research is needed on the natural variability of burn patterns...''
This study addresses that need in two ways:
\begin{enumerate}
\item Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires
\item Examining the capability of using numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench scale and full scale fire experiments conducted in this study.
\end{enumerate}
This paper highlights the many uncertainties involved in what appear to be simple fire experiments.  A variety of uncertainties related to the composition of the fuels, measurement methods, and analysis techniques just to list a few of the areas considered.  The objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  While the variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, 11~\%, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.  The patterns generated by the polyurethane foam fire, had more variability than the natural gas and the gasoline fire patterns.  This is likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.  The maximum fire pattern heights generated from the natural gas and gasoline fires were shown to have uncertainties of 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits.  The comparison of the fire pattern heights and the mean flame height demonstrated that the ``steady state'' natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.

@@

\acknowledgments{
I would like to thank Prof.~Fleishmann, etc.

This work was partially funded by the National Institute of Justice (NIJ) through an interagency agreement with the NIST Office of Law Enforcement Standards (OLES). I would like to thank Susan Ballou of OLES for her support of this project. This work was also partially supported by the U.S. Fire Administration.

The would also like to thank Roger Nokes for his development and continued assistance with Streams and to the New Zealand Fire Commission for their financial support to the University of Canterbury

Finally, my family...
}

@@

Fire investigation is a key element in the fire prevention system that generates data to enable the reduction of fire losses.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus further reduce the losses from fires.  Data such as the room of fire origin, the first item ignited and ultimately the cause of the fire is information that is critical to understanding and reducing the number of fires.  For example, the identification of products that are involved in ignitions due to a design flaw, in many cases result in the U.S. Consumer Product Safety Commission or a manufacturer issuing a product recall.
In some cases, the fire investigation determines that the fire was intentional.  An intentional fire as defined in the National Fire Incident Reporting System (NFIRS)  are those fires that are deliberately set and include fires that result from deliberate misuse of a heat source, fires of an incendiary nature (arson), as well as controlled burn fires, such as crop clearing, that required fire service intervention~\cite{Campbell:2014}.  An incendiary fire, as defined by the National Fire Protection Association (NFPA) Guide for Fire \& Explosion Investigations, commonly referred to as NFPA~921, is ``a fire that is deliberately set with the intent to cause a fire to occur in an area where the fire should not be.''~\cite{NFPA_921}.

@@

@@

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns~\cite{Shanley:1997,Putorti:1997}. A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s) and compartmentalization, to name a few. Given the limited number of experiments in the literature and the large number of variables, it has been difficult to fully understand a cause and effect relationship between the fire scenarios and the resulting patterns.  Therefore, fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs). These spread sheets incorporate references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study because an insufficient amount of reliable data is available to conduct validation.  Two correlations that could not be validated were for flame height and time to ignition.

@@

Given that fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount or research and analysis tools directly related to this practice.  The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA_921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a ``guide'' as opposed to a ``standard,'' it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

>>>>

\dedication{To my wife and family}

\abstract{
Uncontrolled fires result in loss of life, property and can create an adverse economic impact on a community.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus reduce the losses from fires.  However, as the result of the review of several arson homicide cases and a review forensics by the U.S. National Academy of Sciences, questions about the lack of the use of science in the practice of fire investigation were raised.  Specifically the National Academy of Sciences indicated that ``... research is needed on the natural variability of burn patterns...''
This study addresses that need in two ways:
\begin{enumerate}
\item Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires
\item Examining the capability of using numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench scale and full scale fire experiments conducted in this study.
\end{enumerate}
This paper highlights the many uncertainties involved in what appear to be simple fire experiments.  A variety of uncertainties related to the composition of the fuels, measurement methods, and analysis techniques just to list a few of the areas considered.  The objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  While the variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, 11~\%, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.  The patterns generated by the polyurethane foam fire, had more variability than the natural gas and the gasoline fire patterns.  This is likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.  The maximum fire pattern heights generated from the natural gas and gasoline fires were shown to have uncertainties of 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits.  The comparison of the fire pattern heights and the mean flame height demonstrated that the ``steady state'' natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.

@@

\acknowledgments{
I would like to thank Prof.~Fleishmann, Dr.~Kevin McGrattan, and Dr.~Craig Weinschenk from the National Institute of Standards and Technology for their guidance, assistance, and support throughout this study.

The full-scale and bench-scale experiments were conducted over a period of several years whenever funding and laboratory availability coincided.  These experiments were conducted with the support of several engineering technicians from the NIST Fire Research Division, especially Roy McLane, Jay McElroy, Laurean DeLauter, and Anthony Chakalis.  Their assistance was essential. Their patience and attention to detail during this study is greatly appreciated

This work was partially funded by the National Institute of Justice (NIJ) through an inter-agency agreement with the NIST Office of Law Enforcement Standards (OLES). I would like to thank Susan Ballou of OLES for her support of this project. This work was also partially supported by the U.S. Fire Administration.

I would also like to thank Roger Nokes for his development and continued assistance with Streams and to the New Zealand Fire Commission for their financial support to the University of Canterbury

Last but not least, I thank my wife, Liz, for her patience and support of my pursuit of higher education throughout our 37 years of marriage.  I would also like to thank my children; Catherine, Stephanie, Michael, Christina, and Veronica for "buying me time" to work on my studies.}

@@

Fire investigation is a key element in the fire prevention system that generates data to enable the reduction of fire losses.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus further reduce the losses from fires.  Data such as the room of fire origin, the first item ignited and ultimately the cause of the fire is information that is critical to understanding and reducing the number of fires.  For example, the identification of products that are involved in ignitions due to a design flaw, in many cases result in the U.S. Consumer Product Safety Commission or a manufacturer issuing a product recall.

In some cases, the fire investigation determines that the fire was intentional.  An intentional fire as defined in the National Fire Incident Reporting System (NFIRS)  are those fires that are deliberately set and include fires that result from deliberate misuse of a heat source, fires of an incendiary nature (arson), as well as controlled burn fires, such as crop clearing, that required fire service intervention~\cite{Campbell:2014}.  An incendiary fire, as defined by the National Fire Protection Association (NFPA) Guide for Fire \& Explosion Investigations, commonly referred to as NFPA~921, is ``a fire that is deliberately set with the intent to cause a fire to occur in an area where the fire should not be''~\cite{NFPA_921}.

@@

The values cited above are believed to be the best available in the United States and perhaps one of the better fire incident databases in the world.  However an investigative report by Icove and Hargrove, indicates that arson fires are underreported to the National Fire Incident Reporting System (NFIRS).  The investigation provided examples from several major cities, such as New York City, where in 2011, 11 arson fires in buildings were reported to NFIRS, while the Fire Department of the City of New York's Bureau of Fire Investigation had recorded 1,347 arson fires in buildings that year.  Similar discrepancies were found in eight other large U.S. cities.  On average, three-fourths of the arsons uncovered by city fire investigators went unreported to NFIRS~\cite{Icove_2014}.

@@

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns~\cite{Shanley:1997,Putorti:1997}. A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s) and compartmentalization, to name a few. Given the limited number of experiments in the literature and the large number of variables, it has been difficult to fully understand a cause and effect relationship between the fire scenarios and the resulting patterns.

Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs). These spread sheets incorporate references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study because an insufficient amount of reliable data is available to conduct validation.  Two correlations that could not be validated were for flame height and time to ignition.

@@

Given that fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount or research and analysis tools directly related to this practice.  Gorbett recently conducted a review of fire pattern research that has been conducted over the last 80 years. The review points out many gaps, especially with how fire patterns are used to determine the area of origin.
A key limitation is the assumption by many of the researchers, or text authors, that investigators are able to visibly assess varying degrees of fire damage and then determine the area of origin~\cite{Gorbett_2015}. Studies have shown that investigators have not been able to use only the location of the fire patterns as means of reliably determining the area of origin, some amount of analysis and fire dynamics knowledge is required~\cite{Carmen_2008},~\cite{Tinsley_2013}.

The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA_921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a ``guide'' as opposed to a ``standard,'' it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.
